\documentclass[letterpaper, openany, 10pt]{article}
\usepackage[headheight=20pt, total={5in, 8in}]{geometry}
%\usepackage{mathpazo}		% this is the 'proper' way to load palladio, but it breaks blackboard fonts in comparison to Vakil's text 
\usepackage{palatino}
\linespread{1.05}         % Palladio needs more leading (space between lines)
\usepackage[T1]{fontenc}
\usepackage{eulervm}	% Math font that goes well with palladio
%\usepackage{tgpagella} % Alternative font
\usepackage[svgnames]{xcolor}

\title{Hilbert 90 Geometry: Remarks on Birationality}
\author{}
\date{\today}

% All other preamble settings
\input{preamble.tex}

\begin{document}

\maketitle

\section{Remarks on Birationality}

Let $L/k$ be a cyclic field extension of degree $n$, with $p \coloneqq \characteristic k$. Let $G \coloneqq \Gal(L/k) = \left\langle \sigma \right\rangle$. Recall the maps
\begin{gather*}
	f: L^\times \to L^x \\
	x \mapsto x/\sigma(x)
\end{gather*}
and
\begin{gather*}
	h: L \times L \to L \\
	(y, z_0) \mapsto \sigma^0(z_0) + y \cdot \sigma(y) \cdot \sigma^2(z_0) + \ldots + y \cdot \sigma(y) \cdot \ldots \cdot \sigma(y)^{n-2} \cdot \sigma^{n-1}(z_0)
\end{gather*}

We know $f$ maps surjectively onto the subset $X \coloneqq \{y \in L \mid N(y) = 1\}$. From Miles' note, we know that $z_0 \in L$ can be fixed so that $\Tr(z_0) \neq 0$ in order for $h(1, z_0) \neq 0$, which guarantees that $h_{z_0} \coloneqq h|_{L \times \{z_0\}}$, thought of as a polynomial endomorphism in the coefficients of $L \cong k^n$, is at least nonzero at $1 \in L$. If we instead regard $h: L \to \Hom_k(L, L)$ as a function of $y$, we can also recall from the note that
\begin{gather*}
	h \circ f: L^\times \to \Hom_k(L, L) \\
	x \mapsto (z_0 \mapsto x \cdot \Tr(\frac{z_0}{x})).
\end{gather*}

Let $1, x_1, \ldots, x_{n-1}$ be a $k$-basis for $L$, and let
\[
	x = a_0 + a_1x_1 + \ldots + x_{n-1}
\]
be an element of $L$ for some coefficients $a_i \in K$. Let $H$ be the hypersurface in $\mathbb{P}_k^n = \Proj k[a_0, \ldots, a_{n-1}, y]$ given by the equation $N(x) = y^n$. By an abuse of notation, let $f$ denote the map $\mathbb{P}_k^{n-1} = \Proj k[a_0, \ldots, a_{n-1}] \to H$ induced by $f$ as referenced above, and let $g$ denote the map $H \to \mathbb{P}_k^{n-1}$ induced by $h_{z_0}$.

In this note, we will build up to the following proposition:

\begin{proposition}
	If $L/k$ as above is either an Artin-Schreier or Kummer extension, then $f$ is a birational morphism with inverse $g$.
\end{proposition}

We first establish some basic facts. Let $x_0, \ldots, x_{n-1}$ denote a $k$-basis for $L$ such that we represent an arbitrary element $x \in L$ by the sum
\[
	x = \sum_{i=0}^{n-1} a_i x_i
\]
for some indeterminates $a_0, \ldots, a_{n-1}$ representing a choice of coefficients in $k$. With this notation, the field norm $N(x) \coloneqq N_{L/k}(x)$ can be regarded as a polynomial in the indeterminates $a_0, \ldots a_{n-1}$. We recall the following theorem:

\begin{theorem}[\cite{flanders}]
	Let $K = k(\theta)$ be a primitive extension of $k$. Then the general norm $N(x)$ (as a polynomial function of the $a_i$) is irreducible in $k[x]$.
\end{theorem}

\begin{lemma}
	The polynomial $1 - N(x)$ is irreducible.
\end{lemma}

\begin{proof}
	By \cite[Theorem 1]{flanders} above, we know that the polynomial $N(x)$ in the $a_i$ is irreducible. We first show that $1-N(x)$ is irreducible over $k$. To show that $1 - N(x)$ is irreducible, it suffices to show that its homogenization in $\mathbb{P}_k^n$, namely $y^n - N(x)$, is irreducible. By \cite[Exercise 4.5.F(c)]{vakil}, one can check the primeness of a homogeneous ideal or element by considering its divisibility by homogeneous elements.

	So, suppose that $y^n - N(x) = pq$ for two homogeneous polynomials $p,q \in k[a_0, \ldots, a_{n-1}, y]$. We can regard each of $p,q$ as polynomials in $y$ with coefficients in $k[a_0, \ldots, a_{n-1}]$. With this view, it is clear that the product of the terms of $p$ and $q$ which do not involve $y$ is just $N(x)$. But by the irreducibility of $N(x)$, this means that one of $p$ or $q$ must be $1$. Then our homogeneity assumption forces one of $p$ or $q$ to be an element of $k$ and therefore a unit, and this finishes.

	To see that $y^n - N(x)$ is geometrically irreducible, we suppose again that there is some $p,q \in \overline{k}[a_0, \ldots, a_{n-1}, y]$ for which $y^n - N(x) = pq$. Let $l/k$ denote the finite field extension generated by the coefficients of $p$ and $q$ over $k$. Then $y^n - N(x)$ is reducible over $l$. But $N(x)$ describes the field norm of the degree $n$ extension of $l$, so this reducibility contradicts the theorem above.
\end{proof}

\begin{corollary}
	The hypersurface $X$ is integral.
\end{corollary}

\begin{lemma}
	The maps $f$ and $g$ are rational. If $k$ is infinite, then $g$ is also dominant.
\end{lemma}

\begin{proof}
	Both $f$ and $g$ are clearly rational -- they are regular over the complement of the simultaneous vanishing loci of their respective coordinates. If $k$ is infinite, then the subset of $k$-points of $\mathbb{P}_k^{n-1}$ is dense. Let $P \subset \mathbb{P}_k^{n-1}$ denote the subset of $k$-points. The computations from Miles' note show that $(g \circ f)|_P = \id_P$. Hence, $g$ surjects onto the $k$-points of $\mathbb{P}_k^{n-1}$, so the image of $g$ is dense in $\mathbb{P}_k^{n-1}$, and this finishes.
\end{proof}

Since $g$ is dominant, we can realize the function field $K(X)$ as an extension of $K(\mathbb{P}_k^{n-1}) = k(t_1, \ldots, t_{n-1})$. By the principal ideal theorem, we also know that $\dim X = n-1$, so $K(X)$ is also a field of transcendence degree $n-1$. It is then a straightforward algebraic exercise to show that the extension $K(X)/K(\mathbb{P}_k^{n-1})$ must be algebraic. 

\begin{claim}
	There is a dense open subset of $\mathbb{P}_k^{n-1}$ over which the following identity holds:
	\[
		g \circ f = \id_{\mathbb{P}_k^{n-1}}.
	\]
\end{claim}

\begin{proof}[Sketch of idea]
	It suffices to show this identity after base-change to the algebraic closure, so we may suppose that $k = \overline{k}$. Since $k$ can be taken to be infinite in this way, the subset of $k$-points, which are precisely the closed points, of $\mathbb{P}_k^{n-1}$ is dense. It would be nice to write some formula for the maps which would give a bijection on these $\overline{k}$ points. If one can show that $X$ is normal, then it might be possible to use Zariski's Main Theorem (e.g.\cite[\href{https://stacks.math.columbia.edu/tag/03GW}{Lemma 03GW}]{stacks-project} and \cite{264216}) to argue that $f$ and $g$ must be birational inverses. 

	Another possibility is working within $k \neq \overline{k}$. Without modifying the existing coordinates, it might be possible to show that the $g \circ f$ and $\id_{\mathbb{P}_k^{n-1}}$ induce the same endomorphism of the $\overline{k}$-points $\mathbb{P}_k^{n-1}(\overline{k})$ to apply the fact referenced \href{https://math.stackexchange.com/q/630465}{here}. One could then immediately apply this to the field extension discussed above to see that $K(X) \cong K(\mathbb{P}_k^{n-1})$, and this would finish.

	Applying these ideas to Artin-Schreier extensions of finite fields would then be a matter of applying descent: since $\SheafHom$ is a sheaf in any Grothendieck topology, we can change base by the faithfully flat map $\Spec k(t) \to \Spec k$ to apply the ideas above which require an infinite base field.
\end{proof}

% Bibliography

\bibliography{note}

\end{document}
